Imposed form in the Early Acheulean? Evidence from Gona, Afar, Ethiopia

### 2023-07-07 manuscript/manuscript.Rmd
introduction updates by Dietz.

---
title: "Imposed form in the Early Acheulean? Evidence from Gona, Afar, Ethiopia"
author:
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA;  dwstout@emory.edu]
- Antoine Muller^[Institute of Archaeology, Mount Scopus, The Hebrew University of Jerusalem, Jerusalem, Israel; antoine.muller@mail.huji.ac.il]
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Michael J. Rogers^[Department of Anthropology, Southern Connecticut State University, New Haven, CT, USA; rogersm1@southernct.edu]
- Sileshi Semaw^[Centro Nacional de Investigación sobre la Evolución Humana (CENIEH), Burgos, Spain; sileshi.semaw@cenieh.es]
abstract: "TBD.\\par \\textbf{Keywords:} Gona; TBD; TBD; TBD; TBD; TBD; TBD"
date: "`r Sys.Date()`"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
editor_options: 
  markdown: 
    wrap: sentence
---

@@

The imposition of intended form on artifacts has long been viewed as a watershed in human cognitive and cultural evolution and is most commonly associated with the emergence of "Large Cutting Tools" (LCTs) in the Early Acheulean [@isaac1976; @kuhnEvolutionPaleolithicTechnologies2020; @holloway1969]. However, this interpretation of Acheulean LCTs as intentionally designed artifacts remains controversial. Alternative proposals range from the possibility that LCTs were unintended by-products of flake production [@moore2016; @noble1996] to the suggestion that their form was "at least partly under genetic control" [@corbey2016]. Even accepting that LCT form was to some extent intended, there is substantial disagreement over the specificity of design. Some analyses have indicated that shape variation in Acheulean handaxes is largely a result of resharpening [@iovita2011; @mcpherron2000] whereas others find form to be unrelated to reduction intensity and more likely to reflect normative expectations of what handaxes should look like [@garcía-medrano2019; @shiptonHandaxeReductionIts2015a; @shipton2020]. Debates about shape of Acheulean LCTs may appear narrowly technical but have broad relevance for evolutionary questions including the origins of human culture [@corbey2016; @tennie2017; @shiptonHandaxeReductionIts2015a], language [@stout2012], teaching [@gärdenfors2017], brain structure [@hecht2015], and cognition [@stout2015; @wynn2016].
To examine these questions, we studied the complete collection of Early Acheulean flaked pieces from 5 sites at Gona and compared them with Oldowan cores form 2 published sites.
By comparing shape variation to typological assignment and measures of flaking intensity, we sought to identify technological patterns that might reveal intent.

There appears to be a broad consensus that refined handaxes and cleavers from the later Acheulean resulted from procedurally elaborate, skill intensive, and socially learned production strategies [@caruana2020; @garcía-medrano2019; @moore2020; @sharon2009; @shiptonEvolutionSocialTransmission2019; @stout2014] although debate over the presence of explicit, culturally transmitted shape preferences continues [@iovita2011; @moore2020; @shipton2020; @wynn2018]. However, there is less agreement regarding the cruder LCTs that are typical of the earliest Acheulean [@beyene2013; @delatorre2018; @diez-martín2015; @lepre2011; @semaw2018] and which continue to occur with variable frequency in later time periods [@mcnabb2015] including in eastern Asia [@li2021].
Although several formal types have been recognized, many workers now see continuum, including the possibility that flake production remained an important or even primary purpose of LCT reduction (Shea).
This questions imposed form.
It is clear that LFB is new but not clear t

Types are questionable, Kuhn on LFB.
Isaac but some on cobbles.
It is even controversial whether asia is "acheulean" Prevailing opinion, but Beyene.
A conservative interpretation of available evidence is that LCT production was guided by a recurring set of functional, ergonomic, and aesthetic design preferences [@wynn2018] with other elements free to vary in response to raw materials, use life, and random population dynamics like drift, bottlenecks, and founder effects [@kuhnEvolutionPaleolithicTechnologies2020; @lycett2016].

@@

@@

@@

Acheulean cobbles are larger and more cylindrical (geologically “rollers”). Shape difference may reflect availability and/or selectivity for flake-able shapes.

>>>>

---
title: "Imposed form in the Early Acheulean? Evidence from Gona, Afar, Ethiopia"
author:
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA;  dwstout@emory.edu]
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Antoine Muller^[Institute of Archaeology, Mount Scopus, The Hebrew University of Jerusalem, Jerusalem, Israel; antoine.muller@mail.huji.ac.il]
- Michael J. Rogers^[Department of Anthropology, Southern Connecticut State University, New Haven, CT, USA; rogersm1@southernct.edu]
- Sileshi Semaw^[Centro Nacional de Investigación sobre la Evolución Humana (CENIEH), Burgos, Spain; sileshi.semaw@cenieh.es]
abstract: "TBD.\\par \\textbf{Keywords:} Gona; TBD; TBD; TBD; TBD; TBD; TBD"
date: "`r Sys.Date()`"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
editor_options: 
  markdown: 
    wrap: sentence
---

@@

The imposition of intended form on artifacts has long been viewed as a watershed in human cognitive and cultural evolution and is most commonly associated with the emergence of "Large Cutting Tools" (LCTs) in the Early Acheulean [@isaac1976; @kuhnEvolutionPaleolithicTechnologies2020; @holloway1969].
However, this interpretation of Acheulean LCTs as intentionally designed artifacts remains controversial.
Alternative proposals range from the possibility that LCTs were unintended by-products of flake production [@moore2016; @noble1996] to the suggestion that their form was "at least partly under genetic control" [@corbey2016].
Even accepting that LCT form was to some extent intended, there is substantial disagreement over the specificity of design.
Some analyses have indicated that shape variation in Acheulean handaxes is largely a result of resharpening [@iovita2011; @mcpherron2000] whereas others find form to be unrelated to reduction intensity and more likely to reflect normative expectations of what handaxes should look like [@garcía-medrano2019; @shiptonHandaxeReductionIts2015a; @shipton2020].
Such debates about shape of Acheulean LCTs may appear narrowly technical but have broad relevance for evolutionary questions including the origins of human culture [@corbey2016; @tennie2017; @shiptonHandaxeReductionIts2015a], language [@stout2012], teaching [@gärdenfors2017], brain structure [@hecht2015], and cognition [@stout2015; @wynn2016].
To examine these questions, we studied the complete collection of Early Acheulean flaked pieces from 5 sites at Gona Project Area and compared them with Oldowan cores form 2 published sites at Gona.
By comparing shape variation to measures of flaking intensity and patterning, we sought to identify technological patterns that might reveal intent.

## Origin of design

There is a broad consensus that refined handaxes and cleavers from the later Acheulean resulted from procedurally elaborate, skill intensive, and socially learned production strategies [@caruana2020; @garcía-medrano2019; @moore2020; @sharon2009; @shiptonEvolutionSocialTransmission2019; @stout2014] although debate over the presence of explicit, culturally transmitted shape preferences continues [@iovita2011; @moore2020; @shipton2020; @wynn2018].
There is much less agreement regarding the less heavily worked and formally standardized LCTs typical of the earliest Acheulean [@beyene2013; @delatorre2018; @diez-martín2015; @lepre2011; @semaw2018].
Such forms continue to occur with variable frequency in later time periods [@mcnabb2015], and may be especially prevalent in eastern Asia [@li2021].
Although formal types have been recognized in the Early Acheulean and are commonly used to describe assemblages, many workers now see a continuum of morphological variation [@duke2021; @kuhn2020; @presnyakova2018]including the possibility that simple flake production remained an important [@shea2010] or even primary [@moore2016] purpose of Early Acheulean large core reduction.

Typologically, LCTs are differentiated from Mode 1 pebble cores on the basis of size (\>10cm) and shape (elongation and flattening) [e.g., @isaac1977].
This consistent production of large, flat, and elongated cores in the Achuelean has long been thought to reflect the pursuit of desired functional and ergonomic properties for hand-held cutting tools [@wynn2018].
Unplanned flaking can sometimes produce cores that fall into the LCT shape range [@moore2016] and this is one possible explanation of the relatively small "protobifaces" that occur in low frequencies in Oldowan assemblages [@isaac1997].
However, the Early Acheulean is clearly distinguished from the Oldowan by the production of larger artifacts necessitating the procurement and exploitation of larger raw material clasts.
Although studies of handaxe variation often focus on shape rather than size, this shift is an important aspect of artifact design with relevance to both production and function.

Production of larger tools was accomplished either through a novel process of detaching and working Large Flake Blanks (LFBs) from boulder cores or simply by using larger cobble and slab cores [@delatorre2018; @isaac1969; @semaw2018].
Both may involve similar flaking "strategies" (e.g., bifacial or multifacial exploitation) to those present in the Oldowan [@duke2021] but require more forceful percussion to detach larger flakes.
This increases the perceptual motor difficulty of the task [@stout2002] and in many cases may have been accomplished using different percussive techniques and supports [@semaw2009].
These new challenges would have increased raw material procurement [@shea2010] and learning costs [@pargeter2019] as well as the risk of serious injury [@gala2023] associated with tool production.
This strongly implies intentional pursuit of offsetting functional benefits related to size increase.
These likely included tool ergonomics and performance [@key2017] as well as flake generation, resharpening, and reuse potential [@shea2010].
Early Acheulean LCT production is thus widely seen as a part of shifting hominin behavioral ecological strategies including novel resources and mobility patterns [@linaresmatás2021; @rogers1994].

The degree of intentional design reflected in the shape of Early Acheulean LCTs is more difficult to determine.
For example, LFB production using a simple "least effort" bifacial/discoidal strategy will tend to generate predominantly elongated (side or end struck) flakes [@toth1982] whether or not this is an intentional design target.
Similarly, the difficulty of flaking relatively spherical cobbles [@toth1982] might bias initial clast selection and subsequent reduction toward flat and elongated shapes even in the absence of explicit design targets.
On the other hand, it has been argued that the shape of Early Acheulean LFBs was intentionally predetermined using core preparation techniques [@delatorre2018] and many researchers perceive efforts at intentional shaping in the organization of flake scars on Early Acheulean handaxes and picks [@beyene2013; @delatorre2018; @diez-martín2015; @duke2021; @lepre2011; @semaw2009].
To date, however, the identification of Early Acheulean shaping has generally relied on qualitative assessment by lithic analysts.
Such assessment may in fact be reliable, but is subject to concerns about potential selectivity, bias, and/or overinterpretation [@davidson2002; @moore2016].

\
This qualitative approach stands in contrast to investigations of Later Achuelean shaping

@@

It is even controversial whether asia is "acheulean" Prevailing opinion, but Beyene.
A conservative interpretation of available evidence is that LCT production was guided by a recurring set of functional, ergonomic, and aesthetic design preferences [@wynn2018] with other elements free to vary in response to raw materials, use life, and random population dynamics like drift, bottlenecks, and founder effects [@kuhnEvolutionPaleolithicTechnologies2020; @lycett2016].

@@

PCA on GM-transformed caliper measures (length, 3 breadth, 3 thickness).
Length is maximum dimension, piece oriented so that Br1\>Br3

Typological and technological attributions considered unreliable.
Data grouped according to context (\~2.5 mya Oldowan sites vs. \~1.5 mya Acheulean sites) and blank form (cobble, flake, indeterminant).

Associations between form and reduction intensity are considered as an indicator of "imposed form." Such form could reflect mental templates and/or biased flaking patterns due to functional or technological constraints

@@

Acheulean cobbles are larger and more cylindrical (geologically "rollers").
Shape difference may reflect availability and/or selectivity for flake-able shapes.
